What is an Ancestral Recombination Graph?

### 2022-01-31 paper.tex

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of non (or
rarely) recombining organisms [refs]. The trees capture what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the trees such as allele
% TODO cite the other papers that preceded ralph2020efficiently here.
frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
disequilibrium~\citep{mcvean2002genealogical} and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

@@

\section*{Ancestral recombination graphs}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
Hudson extended the coalescent to include
recombination~\citep{hudson1983properties,hudson1990gene}
and the model has proved to be especially powerful basis for
simulations~\citep{hudson2002generating,kelleher2016efficient,kelleher2020coalescent,
baumdicker2021efficient}.
Hudson's algorithm is concerned with simulating the process of
the coalescent with recombination as efficiently as possible, and operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time. The details of Hudson's algorithm and how the
results of the ancestral process are represented are critical to later
discussions, and so we will explain the process in some detail.
See~\cite{kelleher2016efficient} for a more detailed
description of the algorithm, and how it may be implemented efficiently.

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time. Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples the lineage is ancestral to over the interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $L$, the process begins with $n$ lineages
of the form $[(0, L, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount and spatial
distribution of ancestral material among lineages.
Let $A$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (L - 1)$ where
\[
\nu = \sum_{x \in A}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |A|$ lineages are present, common ancestor events
occur at rate $k(k - 1) / 2$. In a common ancestor event two lineages
are chosed uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancesty for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

The Ancestral Recombination Graph (ARG) was later introduced
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process

NOTES for continuing:

\begin{itemize}
\item The ARG process does not track ancestral material. Recombination events
occur at rate $k \rho / 2$, and we don't complete until there's only one
lineage left. That means the process continues for an exponentially long time.
\item Discuss the way in which the re events work.
\item Discuss the rules for pulling out the trees. This is equivalent to doing
the interval logic above in Hudson's algorithm.
\end{itemize}

\section*{Ancestral graphs and stochastic processes}

% See https://github.com/tskit-dev/what-is-an-arg-paper/discussions/13
% for discussions on this section

The simplest prominent example of genetic ancestry generated by a stochastic process
is the coalescent process \citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}, which describes the random
ancestral tree at one non-recombining locus obtained by sampling a relatively
small number of individuals at random from a large population.

In a recombining organism, the genetic ancestry of a longer segment of DNA is
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.
% jk-note: could nitpick this in that diamonds don't generate tree
% "changes" as such. Maybe tweak this on the next round of revision?

% jk-note: this isn't quite right I think, as it's muddling the
% distinction between a graph stochastic process
% definition with the graph as the encoding of the marginal trees.
There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.

Similar graph encodings have also been introduced to model other mechanisms
which change the local ancestral tree along the genome, such as gene
conversion~\citep{wiuf2000coalescent} and
horizontal gene transfer~\citep{baumdicker2014infinitely}.

@@

\section*{The ARG as a structure}

@@

>>>>

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of non (or
rarely) recombining organisms [refs]. The tree captures what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the trees such as allele
% TODO cite the other papers that preceded ralph2020efficiently here.
frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
disequilibrium~\citep{mcvean2002genealogical} and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

@@

\section*{Ancestral graphs and stochastic processes}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
Hudson extended the coalescent to include
recombination~\citep{hudson1983properties,hudson1990gene}
and the model has proved to be an especially powerful basis for
simulations~\citep{hudson2002generating,baumdicker2021efficient}.
Hudson's algorithm is concerned with simulating the process of
the coalescent with recombination as efficiently as possible, and operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time. The details of Hudson's algorithm and how the
results of the ancestral process are represented are critical to later
discussions, and so we will explain the process in some detail.
See~\cite{kelleher2016efficient} for a more detailed
description of the algorithm, and how it may be implemented efficiently.

\begin{figure}
\caption{\label{fig-sim-statespace} An illustration of the state space
of Hudson's algorithm (A) and the big ARG (B) at a given instant. The
state maintained by Hudson's algorithm is considerably more complicated,
but is guaranteed to only simulate events corresponding to genetic
ancestors of the sample.
}
\end{figure}

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time.
% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples the lineage is ancestral to over the interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $m$, the process begins with $n$ lineages
of the form $[(0, m, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount and spatial
distribution of ancestral material among lineages.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |L|$ lineages are present, common ancestor events
occur at rate $\binom{k}{2}$. In a common ancestor event two lineages
are chosed uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

% NOTE: it would probably be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
The Ancestral Recombination Graph (ARG) was later introduced
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,
griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process. We imagine a graph in which
each edge corresponds to an extant lineage and nodes are events in the
process (the initial $n$ leaf nodes are ``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is simply $k \rho (m - 1)$ in the original ARG
formulation. At a recombination event we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMCRA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

The state-space of the ARG is much simpler than Hudson's algorithm, which
greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the ARG back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the complete ancestry of a sample
in Hudson's algorithm is
$O(\rho^2)$~\citep[Eq.~5.10]{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the vast majority of the events that occur in the ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.

This original formulation which does not keep track of the distribution
of ancestral material among lineages is the so-called ``Big'' ARG.
The ``little'' ARG is the subset of the Big ARG traversed by
Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG]

Similar graph encodings have also been introduced to model other mechanisms
which change the local ancestral tree along the genome, such as gene
conversion~\citep{wiuf2000coalescent} and
horizontal gene transfer~\citep{baumdicker2014infinitely}.
[Mention the
SMC~\citep{mcvean2005approximating,marjoram2006fast}
and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

@@

\section*{The ARG as a data structure}

The first method to explicitly produce an ARG as output is SHRUB developed
by \citet{song2005efficient}.

[WIP jk]

@@

\citep{camara2016inference} is another type of parsimony method.
